---
title: "Running Nicheformer foundational model"
author: ""
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  This package takes any Bioconductor-based object (SCE or SPE) with gene expression counts and Ensembl gene identifiers, and computes embeddings using Nicheformer (theislab/Nicheformer).
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
vignette: >
  %\VignetteIndexEntry{Running Nicheformer foundational model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

Load the packages needed to run `Nicheformer` in R.

```{r}
library(torch)
library(anndataR)
library(SingleCellExperiment)
library(nicheR)
```

# Loading data

Here, we use a subsetted Xenium human colon experiment with 200 cells and 422 genes from [Oliveira et al.](https://www.nature.com/articles/s41588-025-02193-3).

```{r}
sce <- read_h5ad(system.file("extdata", "spe_converted_subset_200_iter_1.h5ad", package = "nicheR"),as ="SingleCellExperiment")
```

# Tokenization

Tokens encode ranked gene positions according to a reference list of Ensembl IDs. Gene expression is first corrected using its technology-specific mean expression. In addition to these gene tokens, `Nicheformer` prepends three context tokens that give the model biological and technical context: modality, species, and technology (assay).

modality:
* dissociated = 3
* spatial = 4

species:
* human = 5
* Homo sapiens = 5
* Mus musculus = 6
* mouse = 6

technology:
* merfish = 7
* MERFISH = 7
* cosmx = 8
* NanoString digital spatial profiling = 8
* Xenium = 9
* 10x 5' v2 = 10
* 10x 3' v3 = 11
* 10x 3' v2 = 12
* 10x 5' v1 = 13
* 10x 3' v1 = 14
* 10x 3' transcription profiling = 15
* 10x transcription profiling = 15
* 10x 5' transcription profiling = 16
* CITE-seq = 17
* Smart-seq v4 = 18

```{r}
data("xenium_means", package="nicheR")
X <- tokenization(sce, xenium_means, specie = 5L, assay_tok = 9L, modality = 4L)
```

# Loading weights

You can install weights directly from Huggingface using the **hfhub** package
which will cache the weights once downloaded.

```{r}
if(!requireNamespace("hfhub"))
  install.packages("hfhub")
library(hfhub)
nf_weights <- hub_download("juanhenao/nicheR", "nicheformer_embedpath.pt")
```

You can also use **BiocFileCache** package to download and cache the file. 

```{r}
if(!requireNamespace("BiocFileCache"))
  BiocManager::install("BiocFileCache")
url <- "https://huggingface.co/juanhenao/nicheR/resolve/main/nicheformer_embedpath.pt"
bfc <- BiocFileCache::BiocFileCache(ask = FALSE)
nf_weights <- BiocFileCache::bfcrpath(bfc, url)
```

The `nicheformer` function builds the model and loads its pre-trained weights, while `run_nicheformer` computes embeddings in batches and supports GPU acceleration.

```{r}
model <- nicheformer(weights = nf_weights)
emb <- run_nicheformer(model, X, batch_size = 8L, device = "cpu")
```

# Embeddings overview

Each cell's embedding has a fixed length of 512, matching the model's hidden dimension, regardless of how many gene tokens were truncated during tokenization.

```{r}
dim(emb)
emb[1:6,1:6]
```

# Session info

```{r}
sessionInfo()
```